Modal Abstractions for Virtualizing Memory Addresses

### 2023-11-16 examples.tex
space hacks, rounding out p2v explanation

\newcommand{\pvmapping}[1]{\mathcal{A}\textsf{PVMappings}(#1)}
\newcommand{\sumapacesfull}[2]{
  \ownGhost\gammaPreds{\authfull{\singletonMap{#1}{#2}}}
}
\begin{figure*}
\[
\begin{array}{l}
  \mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)\stackrel{\triangle}{=} \textsf{ASpace\_Lookup}(\ptablestore,\Xi,m) \ast \\
  \qquad\qquad\qquad \bigast{(\vaddr, \textsf{paddr})\in \ptablestore}{\exists\;(\textsf{l4e, l3e, l2e, l1e, paddr})\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\vaddr\textsf{, l4e, l3e, l2e, l1e, paddr})} \\
  \qquad\qquad\qquad \bigast{(\paddr) \in \Xi}{\exists\; (\textsf{qfrac, q, val,}\vaddr) \ldotp \ulcorner \vaddr = \paddr + \textsf{KERNELBASE} \urcorner \ast  \underbrace{\sumwalkabs\vaddr\qfrac\paddr }_\text{Ghost translation} \ast \\
\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad    \underbrace{\paddr \mapsto_{\mathsf{p}}\{\textsf{qfrac}\}\; \vale}_\text{Physical location} \ast \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner }\\
  \textsf{ where } \\
   \textsf{ASpace\_Lookup}_{\textsf{id}}(\ptablestore,m) \stackrel{\triangle}{=} \lambda\textsf{ cr3val} \ldotp \; \exists \gammaPred \; \ldotp \ulcorner m \; !!\; \textsf{cr3val} = \textsf{Some } \gammaPred \urcorner \ast
   \ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}} \ast \\ \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad \ownGhostpv\gammaPred{\authfull{\pvmapping\Xi}}

\end{array}
\]
\caption{Global Address-Space Invariant in Figure \ref{fig:peraspaceinvariant} extended with a ghost map bookkeeping identity mappings }
  \label{fig:peraspaceinvariant_with_p2v_extension}
  \end{figure*}

\begin{figure}\footnotesize
\begin{lstlisting}[mathescape]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ulcorner \textsf{KERNELBASE} + \paddr > 0 \urcorner \ast \texttt{rbx}\mapsto_{\textsf{r}} \paddr \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}$
add rbx, KERNELBASE ;; shift physical address by identity mapping constant
$\specline{\exists \qfraczero; \qfrac,\textsf{ w, val}, \textsf{P} \ast \texttt{rbx}\mapsto_{\textsf{r}} \paddr + \textsf{KERNELBASE} }_{\rtv}$
$\specline{\bigast{(\vaddr, \paddr)\in \ptablestore}{\exists\;(\textsf{l4e, l3e, l2e, l1e, } \paddr)\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\vaddr\textsf{, l4e, l3e, l2e, l1e,} \paddr)}}_{\rtv}$
$\specline{\mathcal{R} -\ast \bigast{(\paddr) \in \Xi}{\exists\; (\textsf{qfrac, q, val, va}) \ldotp \ulcorner \textsf{va} = \paddr + \textsf{KERNELBASE} \urcorner \ast \sumwalkabs\vaddr\qfrac\paddr  \ast \paddr \mapsto_{\mathsf{p}}\{\textsf{qfrac}\} \; \vale\ast  \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner}}_{\rtv}$
$\specline{ \ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}} \ast \ownGhostpv\gammaPred{\authfull{\pvmapping\Xi}}}$
$\specline{ \sumapaces\rtv\delta \ast \sumpv\paddr\qfraczero\true \ast \textsf{P}}_{\rtv}$
$\specline{\vaddr \mapsto_{\texttt{vpte}} \{\textsf{q}, \textsf{qfrac}\} \paddr \;(\texttt{val})}$
\end{lstlisting}
\todo[inline]{Ismail: Should $\delta$ be something else? $\delta$ was used for $\theta$, we need a different gname for $\Xi$.}
\caption{Converting a physical address of a PTE to a virtual address (without instruction pointer or flag updates).}
\label{fig:p2v}
\end{figure}

Work to verify the full page table walk is ongoing, but we have verified the key step for a small x86-64 kernel, which is the physical-to-virtual
conversion (often appearing as a macro \texttt{P2V} in C source code). In our small kernel, as in larger kernels, \texttt{P2V} is actually just addition,
though the Hoare triple we verify for this add instruction is rather complex, with the specification shown in Figure \ref{fig:p2v}.
Beyond a frame ($P$), the precondition requires the current address space invariant, a check that the addition will not overflow, the physical address to convert
(in \lstinline|rbx|), knowledge that the current address space is valid, and knowledge that the physical address is in a range dedicated
to page table entries (as this physical-to-virtual conversion is only guaranteed for page table memory). The latter uses an additional ghost set added to the per-address-space
invariant per Figure \ref{fig:peraspaceinvariant_with_p2v_extension}. The extended invariant owns, for each memory location dedicated to page tables, some fraction of
that physical memory, and also guarantees the offset is the correct calculation.\todo{Colin needs a second pass here}
This is a simple piece of code whose functionality is critical and whose correctness is highly non-trivial. No prior work engages with this problem.

>>>>

\newcommand{\pvmapping}[1]{\mathcal{A}\textsf{PVMappings}(#1)}
\newcommand{\sumapacesfull}[2]{
  \ownGhost\gammaPreds{\authfull{\singletonMap{#1}{#2}}}
}
\begin{figure*}
\footnotesize
\[
\begin{array}{l}
  \mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)\stackrel{\triangle}{=}\\ \textsf{ASpace\_Lookup}(\ptablestore,\Xi,m) \ast 
  \bigast{(\vaddr, \textsf{paddr})\in \ptablestore}{\exists\;(\textsf{l4e, l3e, l2e, l1e, paddr})\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\vaddr\textsf{, l4e, l3e, l2e, l1e, paddr})} \\
  \bigast{(\paddr) \in \Xi}{\exists\; (\textsf{qfrac, q, val,}\vaddr) \ldotp \ulcorner \vaddr = \paddr + \textsf{OFF} \urcorner \ast  \underbrace{\sumwalkabs\vaddr\qfrac\paddr }_\text{Ghost translation} \ast 
% \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad    
\underbrace{\paddr \mapsto_{\mathsf{p}}\{\textsf{qfrac}\}\; \vale}_\text{Physical location} \ast \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner }\\
  \textsf{ where } \\
   \textsf{ASpace\_Lookup}_{\textsf{id}}(\ptablestore,m) \stackrel{\triangle}{=} \lambda\textsf{ cr3val} \ldotp \; \exists \gammaPred \; \ldotp \ulcorner m \; !!\; \textsf{cr3val} = \textsf{Some } \gammaPred \urcorner \ast
   \ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}} \ast  \ownGhostpv\gammaPred{\authfull{\pvmapping\Xi}}

\end{array}
\]
\vspace{-1em}
\caption{Global Address-Space Invariant in Figure \ref{fig:peraspaceinvariant} extended with a ghost map bookkeeping identity mappings }
  \label{fig:peraspaceinvariant_with_p2v_extension}
  \end{figure*}

\begin{figure}\footnotesize
\begin{lstlisting}[mathescape]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ulcorner \textsf{OFF} + \paddr > 0 \urcorner \ast \texttt{rbx}\mapsto_{\textsf{r}} \paddr \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}$
add rbx, OFF ;; shift physical address by identity mapping constant OFF
$\specline{\textsf{P} \ast \exists \qfraczero; \qfrac,\textsf{ w, val},  \texttt{rbx}\mapsto_{\textsf{r}} \paddr + \textsf{OFF} \ast \sumapaces\rtv\delta \ast \sumpv\paddr\qfrac\true \ast \vaddr \mapsto_{\texttt{vpte}} \{\textsf{q}, \textsf{qfrac}\} \paddr \;(\texttt{val})}$
$\specline{\ulcorner \vaddr = \paddr + \textsf{OFF} \urcorner \ast \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \ast \sumwalkabs\vaddr\qfrac\paddr \ast (R \wand \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)) }_{\rtv}$
\end{lstlisting}
\vspace{-1em}
\caption{Converting a physical address of a PTE to a virtual address (w/o instruction pointer or flag updates). $R$ abbreviates the resources
associated with \paddr\ by $\Xi$ in Figure \ref{fig:peraspaceinvariant_with_p2v_extension}.}
\label{fig:p2v}
\end{figure}

Work to verify the full page table walk is ongoing, but we have verified the critical step for a small x86-64 kernel, which is the physical-to-virtual
conversion (often appearing as a macro \texttt{P2V} in C source code). In our small kernel, as in larger kernels, \texttt{P2V} is actually just addition,
but the correctness of this simple instruction is quite subtle, as shown in Figure \ref{fig:p2v}.
Beyond a frame ($P$), the precondition requires the current address space invariant, a check that the addition will not overflow, the physical address to convert
(in \lstinline|rbx|), knowledge that the current address space is valid, and knowledge that the physical address is in a range dedicated
to page table entries (as this physical-to-virtual conversion is only guaranteed for page table memory). 
The latter uses an additional ghost set added to the per-address-space
invariant per Figure \ref{fig:peraspaceinvariant_with_p2v_extension}. The extended invariant owns, for each memory location dedicated to page tables, some fraction of
that physical memory, and also guarantees the offset is the correct calculation.
Subtly, the operation ensures the relevant table entry is readable, but the exact portion of ownership returned must be determined by inspecting the valid bit
of the value in memory --- so full ownership is returned only for unused entries.
This is a simple piece of code whose functionality is critical and whose correctness is highly non-trivial. No prior work engages with this problem.
